---
title: "Cyclistic"
author: "Jorge Gómez"
date: "2023-03-10"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Desde la fundanción de Cyclistic la estrategia se ha basado en el amplio número de segmentos, usando como sistema para adquirir más clientes los diferentes planes de precio. En estudios anteriores se concluyó que los usuarios con membresía anual son más redituables que los usuarios casuales, por lo que la directora de marketing, Lily Moreno, ha decidido que enfocarse en adquirir más usuarios de este tipo, enfocándose en que los **[usuarios casuales adquieran su membresía anual]{.underline}**.

Se ha decidido que el programa de marketing futuro debe responder las siguientes 3 preguntas:

- **How do annual members and casual riders use Cyclistic bikes differently?**

- **Why would casual riders buy Cyclist annual memberships?**

- **How can Cyclistic use digital media to influence casual riders to become members?**

## Data

Data was provided by *Motivate International Inc.*. It is formatted in CSV files, by month, we are taking 12 months from 2022. Data is stored in our servers.
Data consist in 13 columns:

- ride_id = col_character(),
- rideable_type = col_character(),
- started_at = col_datetime(format = ""),
- ended_at = col_datetime(format = ""),
- start_station_name = col_character(),
- start_station_id = col_character(),
- end_station_name = col_character(),
- end_station_id = col_character(),
- start_lat = col_double(),
- start_lng = col_double(),
- end_lat = col_double(),
- end_lng = col_double(),
- member_casual = col_character()

We will check for empty or null values and de dataset

```{r warning=FALSE}
# indluding libraries
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(forcats)
```

```{r lists-of-files}
#reading files
list_of_files <- list.files(path = "../Data/",
                            recursive = TRUE,
                            pattern = "\\.csv$",
                            full.names = TRUE)
all_trips <- readr::read_csv(list_of_files, id = "file_name")
weekday_names <- c("Sun","Mon", "Tue", "Wen", "Thu", "Fri", "Sat", "Sun")
```


Check for NA values in dataset
``` {r check for NA values}
filter(all_trips, is.na(ended_at) | is.na(started_at) | is.na(rideable_type) | is.na(member_casual))
```



Number of each type of user

```{r count-of-members}
users <- all_trips %>%
  count( member_casual)
print(users)
```

Adding duration per trip 
```{r adding-trip-duration-to-df}
alltripwithDuration <- select(all_trips, ride_id, rideable_type, started_at, ended_at, member_casual)
alltripwithDuration <- mutate(alltripwithDuration, durationSecs = ended_at - started_at)
alltripwithDuration <- mutate(alltripwithDuration, durationMin = alltripwithDuration$durationSecs/60)
```

Boxplot of Trips per weekday
```{r adding-weekday-to-df}

##RUN TO SUMMARIZE STATS FOR VARIABLE
tripsWeekday <- select(alltripwithDuration, started_at, ended_at,  member_casual, rideable_type, durationSecs, durationMin) 
tripsWeekday <- tripsWeekday %>%
  mutate(WeekdayName = wday(started_at, label = TRUE), Weekday = wday(started_at, locale =  Sys.getlocale("LC_TIME")), Duration = ended_at - started_at, DurationInMin = Duration/60)
```


Duration of trips per WeekDay
```{r summary-stats-boxplot}
tripsWeekday %>%
    ggplot() +
    geom_boxplot(mapping = aes(x = WeekdayName, y = DurationInMin))
```

In this boxplot, we can see how are a lot of outliers, even negative outliers (which does not make sense). Let's take an example of 1 of them that lasted for 1000 min, that would be 16 hours!


```{r removing-negative-durations}
#Remove negative durations
  tripsWeekday <- tripsWeekday %>%
    filter(started_at < ended_at)
```

```{r getting-iqr}
#Getting IQR 25 & 75:
groupedByDays <-tripsWeekday%>% group_by(WeekdayName)
Qtls <- tapply(tripsWeekday$DurationInMin, tripsWeekday$WeekdayName , quantile)

Q1s <- sapply(1:7, function(i) Qtls[[i]][2])
Q3s <- sapply(1:7, function(i) Qtls[[i]][4])
```

```{r getting-iqr-limits}
#Getting IQR Limits
IQRs <- tapply(tripsWeekday$DurationInMin, tripsWeekday$WeekdayName, IQR)
IQRs

lower <- Q1s - 1.5 * Q1s
upper <- Q3s + 1.5 * Q3s
```

```{r removing-outliers}
tripswDay <- split(tripsWeekday, tripsWeekday$WeekdayName)

lower[1]
upper[2]

noOutlier <- NULL
for (i in 1:7){
  out <- subset(tripswDay[[i]], (tripswDay[[i]]$DurationInMin > lower[i]) & (tripswDay[[i]]$DurationInMin < upper[i]))
 noOutlier <- rbind(noOutlier, out) 
}
```

Trips Without outliers
```{r boxplot-without-outliers}

noOutlier %>%
  ggplot() +
  geom_boxplot(mapping = aes(x = WeekdayName, y = DurationInMin), outlier.shape = NA )

```

- Now, the boxplot is clear from positive and negative outliers, and we can see the area of the box.

```{r number-trip-per-month}
tripsPerMonth <- noOutlier %>%
  count(month(started_at))
tripsPerMonth$`month(started_at)` <- month.abb[tripsPerMonth$`month(started_at)`]

tripsPerMonth %>%
  ggplot(mapping =  aes(x =`month(started_at)`, y = n)) +
  geom_bar(stat='identity') + 
  scale_x_discrete(limits = month.abb)+
  scale_y_continuous(labels = scales::comma)+
  labs(
    title = 'Number of trips per month',
    x = 'Month',
    y = 'Number of trips'
  )

```

-Weather and vacations is a direct factor for the number of rides.

```{r Trips per month by costumer}
#Trips per month by costumer
tripsPerMonthByConstumer <- noOutlier %>%
  group_by(member_casual, month(started_at)) %>%
  count(member_casual)

tripsPerMonthByConstumer <- tripsPerMonthByConstumer %>%
  mutate(monthName = month.abb[`month(started_at)`])

ggplot(data = tripsPerMonthByConstumer) +
  geom_bar(mapping = aes(x = fct_inorder(monthName), y = n,fill = member_casual), stat='identity', position = "dodge") +
  labs( title = "Number of rides each month per type of user",
        x = "Month",
        y = "# of rides"
  )+
  scale_y_continuous(labels = scales::comma)
```

- It is very notorious how members are always at the top of number of rides, but on wynter, casual rides don't even reach half of the member usage.

```{r number of trips per weekday and costumer}
#Trips per weekday and costumer
countOfTripsPerDayCostumer <- noOutlier %>%
  count(Weekday, member_casual) %>%
  mutate(Weekday = weekday_names[Weekday])

countOfTripsPerDayCostumer %>%
  ggplot(mapping = aes(x =fct_inorder(Weekday), y = n, fill = member_casual)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(
    x = "Weekday",
    y = "# of rides",
    
  )+
  scale_y_continuous(labels = scales::comma)+
  scale_fill_discrete(name ="Customer")

```

- On weekends the number of rides are almost even for the 2 types of users. 

```{r AVG ride duration per costumer}
#AVG ride duration and costumer

avgDurationByCustomer <- noOutlier %>%
  group_by(member_casual) %>%
  summarize(mean(durationMin))

avgDurationByCustomer %>%
ggplot(mapping = aes(x = member_casual, y = `mean(durationMin)`), color = member_casual ) +
  geom_bar(stat = 'identity')+
  labs(
    x = 'Member',
    y = 'AVG Duration',
    title = 'Average duration'
  )

tripDurationWeekAndCostumer <- noOutlier %>%
  group_by(Weekday, WeekdayName, member_casual)

tripsWeekday %>%
  group_by(Weekday, WeekdayName, member_casual) %>%
  summarize(avgDuration = mean(Duration)/60, .groups = 'drop') %>%
  ggplot(mapping = aes(x = WeekdayName, y = avgDuration, fill = member_casual)) + 
  geom_bar(stat = 'identity', position = 'dodge') + 
  labs(
    title = 'Avg ride duration per day by Costumer',
    y = 'Min'
  )

```

Average duration per trip by member its very linear, showing that it is not for recreational purposes.

```{r rideable-type-per-costumer}
tripsPerRideAndCustomer <- noOutlier %>%
  count(rideable_type, member_casual)
ggplot(data = tripsPerRideAndCustomer) + 
  geom_bar(mapping = aes(x = rideable_type, y = n, fill = member_casual), stat= 'identity', position = "dodge")+
  labs(
    title = 'Type of ride per costumer',
    x = 'Rideable type',
    y = 'Number of users'
  )
```


```{r count-of-trips-per-hour-customer}
repeatedhours <- noOutlier %>%
  count(hour(started_at), member_casual)

repeatedhours

repeatedhours%>%
  ggplot(mapping = aes(x = `hour(started_at)` , y=n,  color = member_casual )) +
  geom_point() +
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  labs(y = '# of rides per hour',
       x = 'Hour of the day'
       )+
  scale_x_continuous(breaks=seq(0,23,by=1))

```

```{r}
#hist Starting time and day

repeatedhours_day <- noOutlier %>%
  count(hour(started_at),  WeekdayName, member_casual)

repeatedhours_day

graph <- repeatedhours_day%>%
  ggplot(mapping = aes(x = `hour(started_at)` , y=n,  color = member_casual )) +
  geom_point() +
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  labs(y = '# of rides per hour',
       x = 'Hour of the day'
  )+
  scale_x_continuous(breaks=seq(0,23,by=2))

graph + facet_wrap(~WeekdayName, scales = 'free')

```

- Member behavior is very similar through weekdays. From 5 A.M.  to  8 A.M. is a tendency of how rides increase, then it stabilize and increases all over again from 3 P.M and reaching its highest peak at 6 P.M.

- Another thing to notice is that on Saturday and Sunday, the tendency of casual and annual members is very similar, this tell us that on weekends annual members use the service as recreational. 


## Final Comments

1) **Members use the service in a *monotonous* way. On weekdays the peak of usage is at 8:00 hours in the morning and 18:00 in the afternoon, what shows us that the bikes are used most likely to go to work or school. On the other hand, average ride duration of casual members last longer, around 14 min vs almost 11 min that lasts for annual members.**

2) **For casual riders to buy the annual membership, a very good deal should be involved, as they use the bikes about only 6 months of the year.**

3) **Cyclistic could show casual riders the benefits of only using the bikes not just as a leisure activity, but on a daily basis.**




